Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-09-03 submission/manuscript.Rmd
fix order of last fig and add in results

We then identified germline gene promoters differentially methylated in wild-type versus _Kdm5c_-KO exEpiLCs (methylKit, q < 0.01, |methylation difference| > 25%, TSS &#177; 500 bp) and found `r nrow(WTvsKO_hypo)` germline promoters were significantly hypomethylated with _Kdm5c_ loss (Figure 6J). Approximately half of germline promoters hypomethylated in _Kdm5c_-KO exEpiLCs are direct targets of KDM5C in EpiLCs (`r nrow(WTvsKO_hypo_5C)` out of `r nrow(WTvsKO_hypo)` hypomethylated DMRs). Promoters that showed the most robust loss of CpGme (low q-values) harbored CGIs (Figure 6J). We then evaluated promoter CpGme at germline genes ectopically transcibed in either _Kdm5c_-KO EpiLCs or within the brain and found promoter CpGme was substantially reduced in about half of germline DEGs (Figure 6K). Significantly hypomethylated promoters included genes consistently dysregulated across multiple _Kdm5c_-KO RNA-seq datasets[], such as _Naa11_ and _D1Pas1_ (methlyation difference = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "qvalue"),3)`) (Figure 6L). Suprisingly, we observed only a modest reduction in CpGme at _Dazl_'s promoter (methlyation difference = `r signif(getWGBSinfo(WTvsKO, "Dazl", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO, "Dazl", "qvalue"),3)`) (Figure 6M). Altogether, these results demonstrate KDM5C is recruited to germline gene CGIs to promote CpGme at germline gene promoters during early embryogenesis, however some loci can compensate for KDM5C loss through other silencing mechansisms, even when retaining H3K4me.

>>>>

We then identified germline gene promoters differentially methylated in wild-type versus _Kdm5c_-KO exEpiLCs (methylKit, q < 0.01, |methylation difference| > 25%, TSS &#177; 500 bp) and found `r nrow(WTvsKO_hypo)` germline promoters were significantly hypomethylated with _Kdm5c_ loss (Figure 6J). Approximately half of germline promoters hypomethylated in _Kdm5c_-KO exEpiLCs are direct targets of KDM5C in EpiLCs (`r nrow(WTvsKO_hypo_5C)` out of `r nrow(WTvsKO_hypo)` hypomethylated DMRs). Promoters that showed the most robust loss of CpGme (low q-values) harbored CGIs (Figure 6J).

We then compared average percent CpGme at germline gene promtoers with or without CGIs. While the degree of CpGme was largely unchanged for non-CGI promoters, CGI promoters on average had a significant reduction in CpGme with KDM5C loss (Figure 6K) (Non-CGI promoters p = 0.0846, CGI promoters p = 0.0081, Mann-Whitney U test). Significantly hypomethylated promoters included genes consistently dysregulated across multiple _Kdm5c_-KO RNA-seq datasets[@scandagliaLossKdm5cCauses2017], such as _Naa11_ and _D1Pas1_ (methlyation difference = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "qvalue"),3)`) (Figure 6L). Suprisingly, we found only a modest reduction in CpGme at _Dazl_'s promoter (methlyation difference = `r signif(getWGBSinfo(WTvsKO, "Dazl", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO, "Dazl", "qvalue"),3)`) (Figure 6M). Altogether, these results demonstrate KDM5C is recruited to germline gene CGIs to promote CpGme at germline gene promoters during early embryogenesis, however some loci can compensate for KDM5C loss through other silencing mechansisms, even when retaining H3K4me.
